---
title: "Cyclistic Analysis"
author: "Akhil Cyrus"
date: "2022-10-18"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Installing packages for Analyisic
```{r eval=FALSE, include=FALSE}
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("scales")

```

###Loading the packages
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
library(scales)
getwd()
```

setwd("/Users/nivea/Documents/R projects/data")

###Uploading datasets
```{r}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

 Viewing sample data 
```{r}
head(q2_2019)
head(q3_2019)
head(q4_2019)
head(q1_2020)
```

# Column name comparing and making it same for all to join the data

```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

#Renaming Columns to make it consistent

```{r}
q2_2019 <- 
  rename(q2_2019,
         ride_id = "01 - Rental Details Rental ID",
         rideable_type = "01 - Rental Details Bike ID",
         started_at = "01 - Rental Details Local Start Time",
         ended_at = "01 - Rental Details Local End Time",
         start_station_name = "03 - Rental Start Station Name",
         start_station_id = "03 - Rental Start Station ID",
         end_station_name = "02 - Rental End Station Name",
         end_station_id = "02 - Rental End Station ID",
         member_casual = "User Type")
```

```{r}
q3_2019 <- 
  rename(q3_2019,
         ride_id = "trip_id",
         rideable_type = "bikeid",
         started_at = "start_time",
         ended_at = "end_time",
         start_station_name = "from_station_name",
         start_station_id = "from_station_id",
         end_station_name = "to_station_name",
         end_station_id = "to_station_id",
         member_casual = "usertype")
```

```{r}
q4_2019 <- 
  rename(q4_2019,
         ride_id = "trip_id",
         rideable_type = "bikeid",
         started_at = "start_time",
         ended_at = "end_time",
         start_station_name = "from_station_name",
         start_station_id = "from_station_id",
         end_station_name = "to_station_name",
         end_station_id = "to_station_id",
         member_casual = "usertype")
```

#checking Data frames for inconsistencies
```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```

#convert ride_id and rideable_type to character
```{r}
q4_2019 <- mutate(q4_2019,ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type))
q3_2019 <- mutate(q3_2019,ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type))
q2_2019 <- mutate(q2_2019,ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type))

```

#Combing all the data frames into one big data frame
```{r}
total_trips <- bind_rows(q2_2019,q3_2019,q4_2019,q1_2020)
```


#Removing Not necessary fields from total_trips
```{r}
total_trips <- total_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng,birthyear,gender,
            "01 - Rental Details Duration In Seconds Uncapped",
            "05 - Member Details Member Birthday Year","Member Gender",
            "tripduration"))

```

#Cleaning and Add data for Analysis

```{r}
colnames(total_trips)
nrow(total_trips)
dim(total_trips)
head(total_trips)
str(total_trips)
summary(total_trips)
```

#reassigning the vales for consistency - making q1_2020 as default
```{r}
total_trips <- total_trips %>%  
  mutate(member_casual = recode(member_casual,"Subscriber"= "member","Customer" = "casual"))


total_trips$date <- as.Date(total_trips$started_at) #The default format is yyyy-mm-dd

total_trips$month <- format(as.Date(total_trips$date), "%m")

total_trips$day <- format(as.Date(total_trips$date), "%d")

total_trips$year <- format(as.Date(total_trips$date), "%Y")

total_trips$day_of_week <- format(as.Date(total_trips$date), "%A")


```

#Adding ride_length to data frame total_trips
```{r}
total_trips$ride_length <- difftime(total_trips$ended_at,total_trips$started_at)

str(total_trips)
```

# converting ride_length to numeric
```{r}
is.factor(total_trips$ride_length)

total_trips$ride_length <- as.numeric(as.character(total_trips$ride_length))

is.numeric(total_trips$ride_length)
```

#Removing Bad Data

### Removing data which are taken out for quality check
### creating a second dataframe

```{r}
total_trips_v2 <- total_trips[!(total_trips$start_station_name == "HQ QR" | 
                                  total_trips$ride_length<0),]
View(total_trips_v2)
```

# Analysing the Data
```{r}
mean(total_trips_v2$ride_length)
median(total_trips_v2$ride_length)
max(total_trips_v2$ride_length)
min(total_trips_v2$ride_length)

summary(total_trips_v2$ride_length)
```


#comapring with customer type

```{r}
aggregate(total_trips_v2$ride_length ~ total_trips_v2$member_casual, FUN = mean)

aggregate(total_trips_v2$ride_length ~ total_trips_v2$member_casual, FUN = median)

aggregate(total_trips_v2$ride_length ~ total_trips_v2$member_casual, FUN = max)

aggregate(total_trips_v2$ride_length ~ total_trips_v2$member_casual, FUN = min)



aggregate(total_trips_v2$ride_length ~ total_trips_v2$member_casual +
            total_trips_v2$day_of_week,FUN = mean)

total_trips_v2$day_of_week <-
  ordered(total_trips_v2$day_of_week, levels=
                                      c("Sunday", "Monday",
                                        "Tuesday", "Wednesday",
                                        "Thursday", "Friday", "Saturday"))

aggregate(total_trips_v2$ride_length ~ total_trips_v2$member_casual +
            total_trips_v2$day_of_week,FUN = mean)


total_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  #groups by usertype and weekday
  summarise(number_of_rides = n() 
      ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

#Vishuvalisation
```{r}
total_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  #groups by usertype and weekday
  summarise(number_of_rides = n() 
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday,y = number_of_rides, fill = member_casual))+
  geom_col(position = "dodge") + scale_y_continuous(labels = comma)
```

# Vishuvalisation for average duration
```{r}
total_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  #groups by usertype and weekday
  summarise(number_of_rides = n() 
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday,y = average_duration, fill = member_casual))+
  geom_col(position = "dodge") + scale_y_continuous(labels = comma)
```

